---
title: "Open research and reproducibility workshop — code example"
author: "Joana Cruz"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# 1. Purpose of this report

This report provides a reproducible example of how to:

- load and manage packages;
- use project-relative directories rather than hard-coded file paths;
- import data safely;
- carry out basic validation checks before analysis;
- produce a descriptive summary table;
- create consistent figures using a reusable function;
- save outputs in a clear and reproducible way;
- use sustainable coding habits to avoid unnecessary reruns and manual rework.

The aim is not only to produce results, but also to make the workflow understandable, reusable, and easier for someone else to run.

::: {.callout-note}
## Why this example matters

This example brings together several practices from Chapter 7:

- clear script structure;
- visible dependencies;
- project-relative paths;
- input checks;
- data validation;
- reusable code;
- saved outputs;
- light automation through R Markdown;
- sustainable coding through reduced manual repetition.
:::

# 2. Setup

```{r setup, include=FALSE}
# Global chunk options control how code and output are displayed in the report.
# Setting these once at the start keeps the document consistent.

knitr::opts_chunk$set(
  echo = TRUE,       # show code in the report
  warning = FALSE,   # hide warnings from the final report
  message = FALSE,   # hide package startup messages
  fig.width = 8,
  fig.height = 6
)
```

# 3. Load packages

```{r load_packages, include=TRUE}
# In reproducible work, it is helpful to state clearly which packages are needed.
# This avoids situations where code only works on one person's computer.
#
# This helper function is useful for teaching because it makes the script easier
# to run on a new machine. For a real research project, a stronger approach is
# to use renv to record exact package versions.

load_or_install <- function(packages) {
  missing_packages <- packages[!packages %in% installed.packages()[, "Package"]]
  
  if (length(missing_packages) > 0) {
    install.packages(missing_packages)
  }
  
  invisible(lapply(packages, library, character.only = TRUE))
}

# These are the packages used in this report.
#
# here         -> creates project-relative file paths
# fs           -> works with folders and files, including creating directories
# readr        -> useful for reading data files; included here as part of tidy workflows
# dplyr        -> data manipulation
# tibble       -> tidy table objects
# ggplot2      -> plotting
# gtsummary    -> descriptive summary tables
# gt           -> formatting and saving tables
# patchwork    -> combining multiple ggplots into one figure
# RColorBrewer -> colour palettes for plots

required_packages <- c(
  "here",
  "fs",
  "readr",
  "dplyr",
  "tibble",
  "ggplot2",
  "gtsummary",
  "gt",
  "patchwork",
  "RColorBrewer"
)

load_or_install(required_packages)
```

# 4. Define project folders

```{r project_folders, include=TRUE}
# One of the most common reproducibility problems is the use of absolute paths,
# for example:
# "C:/Users/name/Documents/project/data/file.csv"
#
# These paths usually only work on one computer.
#
# The 'here' package solves this by building paths relative to the project folder.
# This means the project can be moved to another computer without rewriting paths,
# as long as the folder structure remains the same.

# In this teaching example, the input datasets are stored in outputs because they
# were generated earlier in the workshop. In a full project, these might be stored
# under data/processed or data/derived.

data_dir <- here::here("outputs")
figures_dir <- here::here("documents", "manuscript", "figures")
tables_dir <- here::here("documents", "manuscript", "tables")

# Create output folders if they do not already exist.
# This avoids errors later when saving figures or tables.
fs::dir_create(figures_dir)
fs::dir_create(tables_dir)

# Print the paths being used so they are visible in the report.
data_dir
figures_dir
tables_dir
```

# 5. Import data

## 5.1 Define the expected files

```{r expected_files, include=TRUE}
# Keeping file names together in one section makes the workflow easier to check
# and update. It also makes the expected inputs explicit.

input_files <- c(
  "dataset_unique.rds",
  "dataset_2018_unique.rds",
  "dataset_2019_unique.rds",
  "dataset_2020_unique.rds",
  "dataset_2021_unique.rds",
  "dataset_2022_unique.rds"
)

input_paths <- file.path(data_dir, input_files)
input_paths
```

## 5.2 Check that files exist

```{r check_files, include=TRUE}
# Before reading files, check that they exist.
# This is a simple but important validation step.
# It is better to fail early with a clear message than to continue and produce
# a confusing error later.

missing_files <- input_paths[!file.exists(input_paths)]

if (length(missing_files) > 0) {
  stop(
    "These required input files are missing:\n",
    paste(missing_files, collapse = "\n")
  )
}
```

## 5.3 Read the main dataset

```{r read_dataset, include=TRUE}
# In this report, the main analyses use the combined dataset.
# We read that file explicitly and give the object a clear, readable name.

nursery_pm25_data <- readRDS(file.path(data_dir, "dataset_unique.rds"))
```

# 6. Basic data validation before analysis

## 6.1 Check that the dataset is not empty

```{r validation, include=TRUE}
# A dataset with zero rows may still load successfully, but it cannot support analysis.

if (nrow(nursery_pm25_data) == 0) {
  stop("The dataset has 0 rows. Analysis cannot continue.")
}

nrow(nursery_pm25_data)
```

## 6.2 Check that required variables exist

```{r variables_present, include=TRUE}
# Define the variables needed for this report.
# Writing them out explicitly makes assumptions visible.

required_vars <- c(
  "year",
  "pm25",
  "pm25_cat",
  "pm25_int",
  "idaci_q",
  "ethnicity",
  "ruc_nm",
  "a_rd_ds_cat"
)

missing_vars <- setdiff(required_vars, names(nursery_pm25_data))

if (length(missing_vars) > 0) {
  stop(
    "The dataset is missing these required variables: ",
    paste(missing_vars, collapse = ", ")
  )
}

required_vars
```

## 6.3 Check variable types

```{r variable_types, include=TRUE}
# Some variables need to be numeric because we summarise or plot them numerically.
# If a variable has unexpectedly become character, that may indicate a data import
# or recoding problem.

if (!is.numeric(nursery_pm25_data$pm25)) {
  stop("Variable 'pm25' must be numeric.")
}

# For this example, year can be numeric, integer, or factor.
# It may be numeric in the raw/processed data and converted to factor for plotting.

if (!is.numeric(nursery_pm25_data$year) &&
    !is.integer(nursery_pm25_data$year) &&
    !is.factor(nursery_pm25_data$year)) {
  stop("Variable 'year' should be numeric, integer, or factor.")
}

# Show the classes of key variables.
sapply(nursery_pm25_data[required_vars], class)
```

## 6.4 Check missingness

```{r missingness, include=TRUE}
# Missing data do not automatically invalidate an analysis,
# but missingness should be understood and documented.

missingness_summary <- tibble::tibble(
  variable = required_vars,
  n_missing = sapply(nursery_pm25_data[required_vars], function(x) sum(is.na(x))),
  pct_missing = round(
    sapply(nursery_pm25_data[required_vars], function(x) mean(is.na(x)) * 100),
    2
  )
)

missingness_summary
```

## 6.5 Check plausible values

```{r plausible_values, include=TRUE}
# Simple range checks can catch silent errors before they affect results.
# The thresholds used here are illustrative. In real work, thresholds should be
# justified using subject knowledge or documented data specifications.

if (any(nursery_pm25_data$pm25 < 0, na.rm = TRUE)) {
  stop("Negative PM2.5 values detected. Check exposure data.")
}

if (any(!nursery_pm25_data$year %in% 2018:2022, na.rm = TRUE)) {
  stop("Unexpected year values detected. Expected years are 2018 to 2022.")
}
```

# 7. Create Table S1

```{r tableS1, include=TRUE}
# gtsummary::tbl_summary() is useful for descriptive tables because it produces
# publication-friendly output with relatively little code.
#
# We specify:
# - the grouping variable ('by = year')
# - the variables to include
# - how continuous variables should be summarised
# - labels that are more readable than raw variable names

table_s1 <- nursery_pm25_data %>%
  gtsummary::tbl_summary(
    by = year,
    include = c(pm25, pm25_cat, pm25_int, idaci_q, ethnicity, ruc_nm, a_rd_ds_cat),
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c(
      "{N_nonmiss}",
      "{mean} ({sd})",
      "{min}",
      "{median}",
      "{max}",
      "{p25} - {p75}"
    ),
    digits = all_continuous() ~ 2,
    label = list(
      pm25 = "PM2.5 annual mean concentration (µg/m³)",
      pm25_cat = "PM2.5 WHO air quality guideline category",
      pm25_int = "PM2.5 WHO air quality interim target category",
      idaci_q = "Income Deprivation Affecting Children Index quintile",
      ethnicity = "Ethnic composition",
      ruc_nm = "Rural–urban area",
      a_rd_ds_cat = "Distance to nearest A road (m)"
    ),
    missing = "ifany"
  ) %>%
  gtsummary::add_n() %>%
  gtsummary::modify_header(label = "**Variable**") %>%
  gtsummary::modify_spanning_header(everything() ~ "**Year**") %>%
  gtsummary::bold_labels()

table_s1
```

## 7.1 Save Table S1

```{r save_table, include=TRUE}
# Saving outputs is part of reproducibility.
# It means the report not only shows the result, but also writes it to a
# predictable location.

table_s1_gt <- gtsummary::as_gt(table_s1)

gt::gtsave(
  data = table_s1_gt,
  filename = file.path(tables_dir, "table_s1_pm25_summary.html")
)
```

# 8. Prepare data for plotting

```{r data_plots, include=TRUE}
# It is often useful to prepare a specific plotting dataset.
# Here we convert variables to factors so that ggplot treats them as categories.

plot_data <- nursery_pm25_data %>%
  dplyr::mutate(
    year = factor(year, levels = 2018:2022),
    idaci_q = factor(idaci_q),
    ethnicity = factor(ethnicity),
    ruc_nm = factor(ruc_nm),
    a_rd_ds_cat = factor(a_rd_ds_cat)
  )
```

# 9. Create a reusable plotting function

```{r plot_function, include=TRUE}
# Rather than repeating almost identical ggplot code several times,
# we define one function that can be reused for different grouping variables.
#
# This improves:
# - readability
# - consistency across plots
# - maintainability
# - sustainability, because it avoids duplicated code and repeated manual edits.

make_pm25_boxplot <- function(data, x_var, x_label, panel_label, x_labels = NULL) {
  
  plot_object <- ggplot2::ggplot(
    data = data,
    ggplot2::aes(
      x = .data[[x_var]],
      y = pm25,
      fill = factor(year)
    )
  ) +
    ggplot2::geom_boxplot(
      outlier.color = "darkgrey",
      color = "lightgrey",
      size = 0.2
    ) +
    ggplot2::scale_fill_manual(
      values = RColorBrewer::brewer.pal(5, "RdYlGn")
    ) +
    ggplot2::geom_hline(
      yintercept = 10,
      linetype = "dotted",
      color = "darkgrey"
    ) +
    ggplot2::labs(
      title = panel_label,
      x = x_label,
      y = "Annual mean concentration of PM2.5 (µg/m³)"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(hjust = -0.1, size = 8),
      axis.text = ggplot2::element_text(size = 8),
      axis.title = ggplot2::element_text(size = 8),
      strip.text = ggplot2::element_text(size = 8),
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
      legend.position = "none"
    ) +
    ggplot2::facet_wrap(~ year, nrow = 1, strip.position = "top")
  
  # Optional relabelling of x-axis categories.
  # This is useful when the coded values in the data are less readable
  # than the labels you want to show in the final figure.
  if (!is.null(x_labels)) {
    plot_object <- plot_object + ggplot2::scale_x_discrete(labels = x_labels)
  }
  
  return(plot_object)
}
```

# 10. Produce Figure 2 panels

```{r figure_panel, include=TRUE}
plot_idaci <- make_pm25_boxplot(
  data = plot_data,
  x_var = "idaci_q",
  x_label = "IDACI quintile",
  panel_label = "a)"
)

plot_ethnicity <- make_pm25_boxplot(
  data = plot_data,
  x_var = "ethnicity",
  x_label = "Ethnicity",
  panel_label = "b)",
  x_labels = c("White", "Other")
)

plot_rural_urban <- make_pm25_boxplot(
  data = plot_data,
  x_var = "ruc_nm",
  x_label = "Rural–urban classification",
  panel_label = "c)",
  x_labels = c("Rural", "Urban")
)

plot_road_distance <- make_pm25_boxplot(
  data = plot_data,
  x_var = "a_rd_ds_cat",
  x_label = "Distance to A road (m)",
  panel_label = "d)",
  x_labels = c("≤200", "200–500", "≥500")
)
```

# 11. Combine and show Figure 2

```{r combine_figure2, include=TRUE}
# patchwork allows several ggplots to be combined in a clear layout.

figure_2 <- (plot_idaci | plot_ethnicity) / (plot_rural_urban | plot_road_distance)

figure_2
```

# 12. Save Figure 2

```{r save_figure, include=TRUE}
figure_2_path <- file.path(figures_dir, "figure_2_pm25_boxplots.jpeg")

ggplot2::ggsave(
  filename = figure_2_path,
  plot = figure_2,
  device = "jpeg",
  dpi = 600,
  width = 8.27,
  height = 11.69,
  units = "in"
)

figure_2_path
```

# 13. Session information

```{r session_info, include=TRUE}
# Recording session information helps others understand the software environment
# used to generate the report.

sessionInfo()
```

# 14. What this workflow demonstrates

This example demonstrates several reproducible coding practices:

- packages are declared at the start;
- folders are defined using project-relative paths;
- required inputs are listed explicitly;
- files are checked before import;
- the dataset is validated before analysis;
- missingness is summarised before interpretation;
- similar plots are generated using a reusable function;
- tables and figures are saved automatically;
- the report itself provides a record of the workflow;
- session information is recorded for transparency.

It also demonstrates sustainable coding: the workflow avoids unnecessary manual repetition, reduces the risk of late errors, and makes rerunning the analysis more reliable.
